Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2012-08-12 paper.tex
minor edits
Added to paper.  Minor code edits.

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e. $P \cap D = \emptyset$, and $L = \abs{P \cup D}$ is the total number of symbols transmitted.

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint probability density function of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}_\pi$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. %If $g(\phi)$ is continuous at $\phi = \tfrac{2\pi}{M}k+\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then 
Then the distribution of $(\sqrt{L}\hat{\lambda}_L, \sqrt{L}\hat{m}_L)$ converges to the bivariate normal with zero mean and covariance
\[
\left( \begin{array}{cc} 
\frac{pA_1 + dA_2}{(p + H d)^2} & 0 \\
0 & pB_1 + dB_2
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = h_2(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}),
\]
\[
A_1 = \expect R_1^2\sin^2(\Phi_1), \qquad A_2 = \var\big( R_1^2\sin^2\fracpart{\Phi_1} \big),
\]
\[
B_1 = \expect R_1^2 \cos^2(\Phi_1) - 1, \;\;\; B_2 = \expect R_1^2 \cos^2\fracpart{\Phi_1} - h_2^2(0).
\]
\end{theorem}

@@

\begin{lemma}~\label{lem:diffatlambdaL}
The derivative of $G_L$ exists, and is equal to zero, at $\hat{\lambda}_L$.  That is,
\[
G_L^\prime(\hat{\lambda}_L) = \frac{d G_L}{d \lambda}(\hat{\lambda}_L) = 0,
\]
\end{lemma}
\begin{IEEEproof}
Observe that 
\[
G_L(\lambda) = \frac{1}{L}\sum_{i \in P} R_i \cos(\lambda + \Phi_i) + \frac{1}{L} \sum_{i \in D} R_i \cos\sfracpart{\lambda + \Phi_i}
\] 
is differentiable everywhere except when $\fracpart{\lambda + \Phi_i} = -\tfrac{\pi}{M}$ for any $i \in D$ where $R_i > 0$.  Let $q_i$ be the smallest number from the interval $[-\tfrac{\pi}{M}, 0]$ such that
\[
 \sin(q_i) > -\frac{\rho_o \sabs{\eta}^2 R_i}{4 L \hat{\rho} \sin(\pi/M)}
\]
where $\eta = e^{-j2\pi/M} - 1$.  Observe that if $R_i > 0$ then $q_i < 0$.  Lemma~\ref{lem:fracpartlambdahatnotpi} shows that,
\[
\sabs{\sfracpart{\hat{\lambda}_L + \Phi_i}} \leq \frac{\pi}{M} + q_i
\]
for all $i \in D$.  Thus, $\sfracpart{\hat{\lambda}_L + \Phi_i} \neq -\tfrac{\pi}{M}$ if $R_i > 0$.  The lemma follows.
\end{IEEEproof}

\begin{lemma}\label{lem:fracpartlambdahatnotpi} Let $q_i, i \in D$ be defined as in Lemma~\ref{lem:diffatlambdaL}.  Then $\sabs{\sfracpart{\hat{\lambda}_L + \Phi_i}} \leq \frac{\pi}{M} + q_i$ for all $i \in D$.
\end{lemma}
\begin{IEEEproof}
Recall that $\hat{d}_i = \hat{d}_i(\hat{\theta}), i \in D$ defined in \eqref{eq:hatdfinxtheta} are the minimsers of the function 
\[
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y(\{d_i, i \in D\})}^2,
\]
defined in~\eqref{eq:SSdatasymbols}. The proof now proceeds by contradiction.  Assume that 
\begin{equation}\label{eq:fraclambassumption}
\sfracpart{\hat{\lambda}_L + \Phi_k} > \frac{\pi}{M} + q_k
\end{equation}
for some $k \in D$.  Recalling the notation $e_k$ defined in Section~\ref{sec:least-squar-estim}, put $r_i = \hat{d}_i e_k$.  We will show that $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$, violating the fact that $\{\hat{d}_i, i \in D\}$ are minimisers of $SS$.  First observe that,
\begin{align*}
Y(\{ r_i, i \in D\}) = \sum_{i \in P} y_ip_i^* + \sum_{i \in D} y_i\hat{r}_i^* = \hat{Y} + \eta y_k\hat{d}_k^*.
\end{align*}
where $\eta = e^{-j2\pi/M} - 1$ and $\hat{Y} = Y(\{ \hat{d}_i, i \in D\})$.  Now,
\begin{align*}
SS&(\{r_i, i \in D\}) \\
&= A - \frac{1}{L} \abs{ Y(\{ r_i, i \in D\}) }^2 \\
&= A - \frac{1}{L} \abs{ \hat{Y} + \eta y_k\hat{d}_k^* }^2 \\
&= A - \frac{1}{L}\abs{\hat{Y}}^2 - \frac{2}{L}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) -  \frac{1}{L}\abs{ \eta y_k}^2\\
&= SS(\{\hat{d}_i, i \in D\}) - D
\end{align*}
where 
\[
D = \frac{2}{L}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) +  \frac{1}{L}\abs{ \eta y_k }^2.
\]
Now $\frac{1}{L}\hat{Y} = \hat{a} = \hat{\rho} e^{j\hat{\theta}}$ from~\eqref{eq:hata} and using~\eqref{eq:yethetadhat},
\[
\frac{1}{L} \hat{Y}^* y_k \hat{d}_k^* = \hat{\rho} \rho_0 R_k e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}},
\]
so that
\begin{equation}\label{eq:Ddefn}
D = 2 \hat{\rho} \rho_0 R_k \Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}}\right) + \frac{1}{L}\abs{\eta}^2\rho_0^2 R_k^2.
\end{equation}
Let $v = \sfracpart{\hat{\lambda}_L + \Phi_k} - \tfrac{\pi}{M}$ so that
\begin{align*}
\eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}} &= (e^{-j2\pi/M} - 1) e^{j \pi/M} e^{jv} \\
&= (e^{-j\pi/M} - e^{j\pi/M}) e^{jv} \\
&= -2 \sin(\tfrac{\pi}{M}) e^{j ( v + \pi/2) }
\end{align*}
and
\begin{align*}
\Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}}\right) &= -2 \sin(\tfrac{\pi}{M}) \cos(v + \tfrac{\pi}{2}) \\
 &= 2 \sin(\tfrac{\pi}{M}) \sin(v)
\end{align*}
Because we assumed~\eqref{eq:fraclambassumption},
\[
0 > v > q_k
\]
and, from the definition of $q_k$,
\[
0 > \sin(v) > -\frac{\rho_o \sabs{\eta}^2 R_i}{4 L \hat{\rho} \sin(\pi/M)}.
\]
Subtituting this into~\eqref{eq:Ddefn} gives $D > 0$, but then $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$, violating the fact that $\{\hat{d}_i, i \in D\}$ are minimisers of $SS$.  Thus,~\eqref{eq:fraclambassumption} is false.

To show that $\sfracpart{\hat{\lambda}_L + \Phi_k} \geq -\frac{\pi}{M} - q_k$ we use contradiction again.  Assume that $\sfracpart{\hat{\lambda}_L + \Phi_k} < -\frac{\pi}{M} - q_k$.  Recalling the notation $e_k^*$ defined in Section~\ref{sec:least-squar-estim}, put $r_i = \hat{d}_i e_k^*$.  Now an argument analagous to that used previously shows that $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$ again.

@@

\begin{lemma}\label{lem:empiricprocc} BLERG
\[
\sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} G_L^\prime(0).
\]
\end{lemma}
\begin{IEEEproof}

>>>>

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 2$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e. $P \cap D = \emptyset$, and $L = \abs{P \cup D}$ is the total number of symbols transmitted.

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint probability density function of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}_\pi$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. %If $g(\phi)$ is continuous at $\phi = \tfrac{2\pi}{M}k+\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then 
Then the distribution of $(\sqrt{L}\hat{\lambda}_L, \sqrt{L}\hat{m}_L)$ converges to the bivariate normal with zero mean and covariance
\[
\left( \begin{array}{cc} 
\frac{pA_1 + dA_2}{(p + H d)^2} & 0 \\
0 & \rho_0^2(pB_1 + dB_2)
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = h_2(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}),
\]
\[
A_1 = \var\big( R_1^2\sin^2(\Phi_1) \big), \qquad A_2 = \var\big( R_1^2\sin^2\fracpart{\Phi_1} \big),
\]
\[
B_1 = \var\big( R_1^2 \cos^2(\Phi_1) \big), \;\;\; B_2 = \var\big( R_1^2 \cos^2\fracpart{\Phi_1} \big).
\]
\end{theorem}

@@

\begin{lemma}~\label{lem:diffatlambdaL}
The derivative of $G_L$ exists, and is equal to zero, at $\hat{\lambda}_L$.  That is,
\[
G_L^\prime(\hat{\lambda}_L) = \frac{d G_L}{d \lambda}(\hat{\lambda}_L) = 0,
\]
\end{lemma}
\begin{IEEEproof}
Observe that 
\[
G_L(\lambda) = \frac{1}{L}\sum_{i \in P} R_i \cos(\lambda + \Phi_i) + \frac{1}{L} \sum_{i \in D} R_i \cos\sfracpart{\lambda + \Phi_i}
\] 
is differentiable everywhere except when $\fracpart{\lambda + \Phi_i} = -\tfrac{\pi}{M}$ for any $i \in D$ with $R_i > 0$.  Let $q_i$ be the smallest number from the interval $[-\tfrac{\pi}{M}, 0]$ such that
\[
 \sin(q_i) > -\frac{\rho_o \sabs{\eta}^2 R_i}{4 L \hat{\rho} \sin(\pi/M)}
\]
where $\eta = e^{-j2\pi/M} - 1$.  Observe that if $R_i > 0$ then $q_i < 0$.  Lemma~\ref{lem:fracpartlambdahatnotpi} shows that,
\[
\sabs{\sfracpart{\hat{\lambda}_L + \Phi_i}} \leq \frac{\pi}{M} + q_i
\]
for all $i \in D$.  Thus, $\sfracpart{\hat{\lambda}_L + \Phi_i} \neq -\tfrac{\pi}{M}$ if $R_i > 0$.  The lemma follows.
\end{IEEEproof}

\begin{lemma}\label{lem:fracpartlambdahatnotpi} Let $q_i, i \in D$ be defined as in Lemma~\ref{lem:diffatlambdaL}.  Then $\sabs{\sfracpart{\hat{\lambda}_L + \Phi_i}} \leq \frac{\pi}{M} + q_i$ for all $i \in D$.
\end{lemma}
\begin{IEEEproof}
Recall that $\{\hat{d}_i = \hat{d}_i(\hat{\theta}), i \in D\}$ defined in \eqref{eq:hatdfinxtheta} are the minimsers of the function 
\[
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\sabs{Y(\{d_i, i \in D\})}^2,
\]
defined in~\eqref{eq:SSdatasymbols}. The proof now proceeds by contradiction.  Assume that 
\begin{equation}\label{eq:fraclambassumption}
\sfracpart{\hat{\lambda}_L + \Phi_k} > \frac{\pi}{M} + q_k
\end{equation}
for some $k \in D$.  Recalling the notation $e_k$ defined in Section~\ref{sec:least-squar-estim}, put $r_i = \hat{d}_i e_k$.  We will show that $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$, violating the fact that $\{\hat{d}_i, i \in D\}$ are minimisers of $SS$.  First observe that,
\begin{align*}
Y(\{ r_i, i \in D\}) = \sum_{i \in P} y_ip_i^* + \sum_{i \in D} y_i\hat{r}_i^* = \hat{Y} + \eta y_k\hat{d}_k^*.
\end{align*}
where $\eta = e^{-j2\pi/M} - 1$ and $\hat{Y} = Y(\{ \hat{d}_i, i \in D\})$.  Now,
\begin{align*}
SS&(\{r_i, i \in D\}) \\
&= A - \frac{1}{L} \sabs{ Y(\{ r_i, i \in D\}) }^2 \\
&= A - \frac{1}{L} \sabs{ \hat{Y} + \eta y_k\hat{d}_k^* }^2 \\
&= A - \frac{1}{L}\sabs{\hat{Y}}^2 - \frac{2}{L}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) -  \frac{1}{L}\sabs{ \eta y_k}^2\\
&= SS(\{\hat{d}_i, i \in D\}) - D
\end{align*}
where 
\[
D = \frac{2}{L}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) +  \frac{1}{L}\abs{ \eta y_k }^2.
\]
Now $\frac{1}{L}\hat{Y} = \hat{a} = \hat{\rho} e^{j\hat{\theta}}$ from~\eqref{eq:hata} and using~\eqref{eq:yethetadhat},
\[
\frac{1}{L} \hat{Y}^* y_k \hat{d}_k^* = \hat{\rho} \rho_0 R_k e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}},
\]
so that
\begin{equation}\label{eq:Ddefn}
D = 2 \hat{\rho} \rho_0 R_k \Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}}\right) + \frac{1}{L}\abs{\eta}^2\rho_0^2 R_k^2.
\end{equation}
Let $v = \sfracpart{\hat{\lambda}_L + \Phi_k} - \tfrac{\pi}{M}$ so that
\begin{align*}
\eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}} &= (e^{-j2\pi/M} - 1) e^{j \pi/M} e^{jv} \\
&= (e^{-j\pi/M} - e^{j\pi/M}) e^{jv} \\
&= -2 \sin(\tfrac{\pi}{M}) e^{j ( v + \pi/2) }
\end{align*}
and
\begin{align*}
\Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}}\right) &= -2 \sin(\tfrac{\pi}{M}) \cos(v + \tfrac{\pi}{2}) \\
 &= 2 \sin(\tfrac{\pi}{M}) \sin(v),
\end{align*}
Because we assumed~\eqref{eq:fraclambassumption}, it follows that $0 > v > q_k$ and, from the definition of $q_k$,
\[
0 > \sin(v) > -\frac{\rho_o \sabs{\eta}^2 R_i}{4 L \hat{\rho} \sin(\pi/M)}.
\]
Subtituting this into~\eqref{eq:Ddefn} gives $D > 0$, but then $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$, violating the fact that $\{\hat{d}_i, i \in D\}$ are minimisers of $SS$.  Thus,~\eqref{eq:fraclambassumption} is false.

To show that $\sfracpart{\hat{\lambda}_L + \Phi_k} \geq -\frac{\pi}{M} - q_k$ we use contradiction again.  Assume that $\sfracpart{\hat{\lambda}_L + \Phi_k} < -\frac{\pi}{M} - q_k$.  Recalling the notation $e_k^*$ defined in Section~\ref{sec:least-squar-estim}, put $r_i = \hat{d}_i e_k^*$.  Now an analagous argument can be used to show that $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$ again.

@@

\begin{lemma}\label{lem:empiricprocc}
\[
\sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} G_L^\prime(0).
\]
\end{lemma}
\begin{IEEEproof}
Put 
\[
Z_L(\lambda) = \sqrt{L}\big(G_L^\prime(\lambda) - \expect G_L^\prime(\lambda) - G_L^\prime(0) + \expect G_L^\prime(0)\big)
\] 
so that
\[
\sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) = Z_L(\hat{\lambda}_L) + \sqrt{L} G_L^\prime(0)
\]
Lemma~? shows that for any $\delta > 0$ and $\nu > 0$, there exists $\epsilon > 0$ such that
\[
\Pr\left\{ \sup_{\sabs{\lambda}<\epsilon}\sabs{Z_L(\lambda)} > \delta  \right\} < \nu
\]
for all positive integers $L$.  Since $\hat{\lambda}_L$ converges almost surely to zero, it follows that for any $\epsilon > 0$,
\[
\lim_{L\rightarrow\infty}\prob\left\{ \sabs{\hat{\lambda}_L} \geq \epsilon \right\} = 0
\] 
and therefore $\prob\{ \sabs{\hat{\lambda}_L} \geq \epsilon \} < \nu$ for all sufficiently large $L$.  Now
\begin{align*}
  \prob&\left\{\sabs{ Z_L(\hat{\lambda}_L) } > \delta \right\} \\
&= \prob\left\{ \sabs{Z_L(\hat{\lambda}_L)} > \delta \;\text{and} \; \sabs{\hat{\lambda}_L} < \epsilon \right\} \\
&\hspace{0.7cm} + \prob\left\{ \sabs{Z_L(\lambda_L)} > \delta  \;\text{and} \; \sabs{\hat{\lambda}_L} \geq \epsilon \right\} \\
&\leq \prob\left\{  \sup_{\lambda < \epsilon} \sabs{ Z_L(\lambda) } > \delta \right\} + \prob\left\{ \sabs{\hat{\lambda}_L} \geq \epsilon \right\} \\
&\leq 2\nu.
\end{align*}
for all sufficiently large $L$.  Since $\nu$ and $\delta$ can be chosen arbitrarily small, it follows that $Z_L(\hat{\lambda}_L)$ converges in probability to zero as $N\rightarrow\infty$.
\end{IEEEproof}
